---
title: "R script for the paper *The cumulative impacts of droughts and N deposition on Norway spruce (Picea abies) in Switzerland based on 38 years of forest monitoring* by Tresch *et* al. 2023"
author: "R codes are written by Simon Tresch , Institute for Applied Plant Biology (IAP)"

date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  bookdown::pdf_document2:
      toc: true
      toc_depth: 3
      number_sections: true
urlcolor: gray
linkcolor: gray
      
bibliography: IAP.bib
link-citations: yes


---

<!-- # ```{r setup, include=FALSE} -->
<!-- # knitr::opts_chunk$set(echo = TRUE) -->
<!-- # ``` -->

<!-- {r, results='asis', echo=FALSE} -->
<!-- with pander   pander(summary(mtcars[, x])) -->


**Important notes**
This   [RMarkdown](https://bookdown.org/yihui/rmarkdown) was created using the R version 4.1.1 [@RCore2021].

This scripts includes data preparation and all statistical models including model diagnostics and plots used for the paper.

All codes are coded by the main authors, based on literature given in the main paper and supplementary material. Some R codes for big plots are not displayed in the PDF output but they can be found in the .RMD file on the GitHub repository:

https://github.com/treschsimon/Tresch_et_al_spruce_paper_R_codes

All data and models can be found on the Mendely data repository:
https://data.mendeley.com/datasets/y8hvtzzkb4/1



In case of any questions please contact the main author.

Please cite the main paper Tresch *et* al. 2022.


```{r RMD setup, include=FALSE}
library(bookdown)

library(pander) # Tabellen Darstellen im R Markdown
panderOptions('knitr.auto.asis', FALSE)

```

# Used packages
```{r setup,echo = T, message=FALSE}
library(Matrix);library(tidyverse) 
```
Used for data manipulation (dplyr) and plotting (ggplot) see [@Wickham2019].

```{r setup2,echo = T, message=FALSE}
library(latex2exp) # LaTex expressions for plots
library(ggpubr) # multiple ggplots
library(xtable) # LaTex table output
library(faraway) # calculation of VIF
library(Hmisc) # correlation of variables
library(corrplot)# matrix of the p-value of the correlation
library(lubridate)# passing to a long format
library(timetk) #correlation-analysis-in-time-series-data



```


```{r setup4,echo = T, message=FALSE}
library(splines) #natural cubic splines
library(brms) # Bayesian multilevel models in R
library(bayesplot) # Visualization in Bayesian workflow

```
Bayesian multilevel models in R were conducted using the brms package from [@Burkner2017] and posterior predictive checks with the help of the bayesplot package [@Gabry2019].

\clearpage
# Descriptive statistics

## Load data

See Material and Methods section of the main paper for more information about the study sites, which are part of the Intercantonal long-term forest observation program running since 38 years [@Braun2021_WDB_EN].



```{r sites, echo = T, message=F,warning=F,fig.align = "center"}
## Load long-term forest monitoring data

mydata<-read_delim("data/final_data_mortality_2022.csv")
head(mydata)

mydata <- mydata %>% 
  mutate(sites=as.factor(sites))

```

## Number of dead trees
```{r descr1, echo = T, message=F,warning=F,fig.align = "center"}

(sum(mydata$trees_living) + sum(mydata$trees_died))
#  trees have been observed 1985-2022.

sum(mydata$trees_died) 
# total trees died from 1985-2022.

mortality_rate <-mydata %>%
  group_by(year_fact) %>%
  summarise(living_sum=sum(trees_living),died_sum=sum(trees_died)) %>%
  mutate(mortality=100*died_sum/(living_sum+died_sum)) %>%
  ungroup() 

# Mortality over all years
(long_term_average_mortality<-mean(mortality_rate$mortality, na.rm=T)) #  0.80 % 1985-2022
```
```{r descr1plot, echo = F, message=F,warning=F,fig.align = "center"}

# mortality per year
mortality_rate_plot <-ggplot(data=mortality_rate,aes(x=year_fact, y=mortality/100)) + 
  geom_line(size=0.5)+
  geom_hline(yintercept=long_term_average_mortality/100, colour="grey20",
             linetype="dotted", size=0.5)+
  annotate("text", x = 1994.3, y = 0.0088, size = 2.2,  colour="grey20", lineheight =
             .9,label = "Long-term avarage 1985-2022: 0.8%",hjust = 1)+
   scale_y_continuous(expand = c(0.015,0),
                      limits = c(0,0.06),labels = scales::percent_format(accuracy = 1),
                      breaks =c(0,0.01,0.02,0.03,0.04,0.05,0.06)) +
   geom_point(size=3, colour="white") + 
  geom_point(size=2) + 
  scale_x_continuous(breaks=seq(1985,2020,5))+
  theme_classic()+
  theme(axis.title=element_text(size=9),axis.ticks = element_line(size=0.1), axis.line=element_line(colour="black"),text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 10, colour = "black"),axis.text.x = element_text(size =10, colour = "black"),axis.line.x=element_line(size = 0.1),axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),panel.border = element_rect(colour = "black",
    fill=NA, size=0.1),panel.grid.minor=element_blank(),panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
    legend.title=element_text(),legend.justification = c(1, 1), legend.position = c(1,1),legend.direction="horizontal",legend.key.height=unit(1,"line"),
    plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position =
          c(0.315,0.99),legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")+
  labs(x="", y= TeX("Annual average mortality rate of Picea abies"))+
  geom_curve(aes(x = 1997, y = 0.013, xend = 1999, yend = 0.01), 
             arrow = arrow(length = unit(0.07, "inch")), size = 0.4, color = "darkblue", curvature = 0.3)+
  annotate("text", x = 1997, y = 0.015, size = 2.8, color = "darkblue", 
           lineheight = .9,label = "Winterstorm 'Lothar'")+
  geom_curve(aes(x = 2002, y = 0.023, xend = 2004, yend = 0.02), arrow = arrow(
    length = unit(0.07, "inch")), size = 0.4, color = "darkred", curvature = 0.3)+
  annotate("text", x = 2002, y = 0.025, size = 2.8, color = "darkred", 
           lineheight = .9,label = "Heat wave 2005")+
geom_curve(aes(x = 2014, y = 0.04, xend = 2016, yend = 0.035), arrow = 
             arrow(length = unit(0.07, "inch")), size = 0.4, color = "darkred", 
           curvature = 0.3)+
  annotate("text", x = 2013.5, y = 0.043, size = 2.8, color = "darkred", 
           lineheight = .9,label = "Cummulative \n heat waves 2015-2020")
mortality_rate_plot


```
## Number of study sites
```{r descr2, echo = T, message=F,warning=F,fig.align = "center"}

sites_per_year <- mydata %>%
  group_by(year_fact)  %>%
  summarise(sites_per_year = n())

sites_per_year_plot <-ggplot(data=sites_per_year,aes(x=year_fact, y=sites_per_year)) + 
  geom_bar(stat="identity", fill="black")+
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,82)) +
  geom_text(aes(label=sites_per_year), vjust=1.6, color="white", size=3.5)+
   scale_x_continuous(breaks=seq(1985,2025,5))+
  theme_classic()+
  labs(x="", y= "Norway spruce sites in the long-term Intercantonal \nForest Observation Program in Switzerland")
sites_per_year_plot

```

## Variable selection and scaling
```{r descr3, echo = T, message=F,warning=F,fig.align = "center"}

# explaining variables
mydata$n_dep_s<-scale(mydata$n_dep, center=TRUE, scale=TRUE) 

# proportion of species mixtures (see for instance Brandl et al. 2020 FEM)
mydata$conifer_s<-scale(mydata$conifer, center=TRUE, scale=TRUE) 

# age
mydata$age_s<-scale(mydata$age, center=TRUE, scale=TRUE) 

# soil variables
mydata$bs_s<-scale(mydata$bs, center=TRUE, scale=TRUE) 

mydata$ph_s<-scale(mydata$ph, center=TRUE, scale=TRUE) 


# Nutrients see Braun2021
mydata$n_s<-scale(mydata$n, center=TRUE, scale=TRUE)

mydata$p_s<-scale(mydata$p, center=TRUE, scale=TRUE)

mydata$k_s<-scale(mydata$k, center=TRUE, scale=TRUE)

mydata$mg_s<-scale(mydata$mg, center=TRUE, scale=TRUE)

mydata$ca_s<-scale(mydata$ca, center=TRUE, scale=TRUE)

mydata$mn_s<-scale(mydata$mn, center=TRUE, scale=TRUE)


# nutrient ratios
mydata$n_k<-mydata$n/mydata$k
mydata$n_k_s<-scale((mydata$n/mydata$k), center=TRUE, scale=TRUE)# n:K
mydata$n_p<-mydata$n/mydata$p
mydata$n_p_s<-scale((mydata$n/mydata$p), center=TRUE, scale=TRUE)# n:p
mydata$n_mg<-mydata$n/mydata$mg
mydata$n_mg_s<-scale((mydata$n/mydata$mg), center=TRUE, scale=TRUE)# n:mg
mydata$n_ca<-mydata$n/mydata$ca
mydata$n_mn_s<-scale((mydata$n/mydata$mn), center=TRUE, scale=TRUE)# n:mg

# Indicators of drought 
# potential drought see Braun et al. 2015 SZF

# actual drought see Braun et al. 2015 SZF
mydata$ETa_ETp_0_s<-scale(mydata$ETa_ETp_0, center=TRUE, scale=TRUE)#lag 0: current year 

mydata$ETa_ETp_1_s<-scale(mydata$ETa_ETp_1, center=TRUE, scale=TRUE)#lag 1: 1 years before

mydata$ETa_ETp_2_s<-scale(mydata$ETa_ETp_2, center=TRUE, scale=TRUE)#lag 2: 2 years before

mydata$ETa_ETp_3_s<-scale(mydata$ETa_ETp_3, center=TRUE, scale=TRUE)#lag 3: 3 years before

mydata$ETa_ETp_4_s<-scale(mydata$ETa_ETp_4, center=TRUE, scale=TRUE)#lag 4: 4 years before

mydata$ETa_ETp_5_s<-scale(mydata$ETa_ETp_5, center=TRUE, scale=TRUE)#lag 5: 5 years before



# Additional explanatory variables
mydata$altitude_s<-scale(mydata$altitude, center=TRUE, scale=TRUE)

mydata$year_s<-scale(mydata$year, center=TRUE, scale=TRUE)# to test remaining year effect 
```
For drought and temperature variables see [@Braun2015c] and for conifer species mixture variable see [@Brandl2020].




## Total N deposition
```{r descr4, echo = F, message=F,warning=F,fig.align = "center", fig.width=10,fig.height=10}

ndep_data <- mydata %>%  mutate(sites = fct_reorder(sites, n_dep, .desc = F))
ndep_data<-ndep_data %>%
  mutate(long_term_median_n_dep = median(n_dep))

n_dep_hist<-ggplot(ndep_data, aes(y=n_dep, x=sites, colour=year_fact)) +
  # geom_boxplot()+
   geom_point()+
  geom_point(data=(ndep_data %>% group_by(sites) %>%
  mutate(n_dep_median_sites = median(n_dep)) %>% unique()), aes(x=sites, y =n_dep_median_sites, fill="black"),colour="black", size=2 )+

  geom_hline(aes(yintercept=long_term_median_n_dep), colour="grey20", linetype="dotted", size=0.5)+
 annotate("text", x = 23, y = 27, size = 2.8,  colour="grey20", lineheight = .9,label = expression(paste("Median 1985-2022: 26 kg N ", ha^-1, " ", a^-1)),hjust = 1)+
  
  scale_fill_manual(values = c("Median per site" = "black")) + 
scale_y_continuous(breaks = c(10, 20,30,40,50,60,70,80))+
  labs(title="",y= expression(paste("Modelled total N deposition (kg N ", ha^-1, " ", a^-1, ")")), x = "Sites (orderd after N deposition)")+
  theme_bw()+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
    theme(legend.position = c(0.09,0.85), axis.title.x=element_text(),axis.text.x=element_text(angle = 45, hjust = 1,size=6),axis.ticks.x=element_blank(),legend.title=element_blank())
n_dep_hist
```
N deposition was modelled according to [@Rihm2019]. All codes for the plot can be found in the .RMD file.
\clearpage
## Lagged effects of drought indicator ETa/ETp
```{r descr5, echo = T, message=F,warning=F,fig.align = "center",fig.width=10,fig.height=10}

mydata$drought_lag <- (mydata$ETa_ETp_0 + mydata$ETa_ETp_1 + 
          mydata$ETa_ETp_2 + mydata$ETa_ETp_3+ mydata$ETa_ETp_4)/5
mydata$drought_lag_s<-scale(mydata$drought_lag, center=TRUE, scale=TRUE)

# see Chapter 3 about the drought effect calculation with the poly-lag model approach
```
```{r descr5plot, echo = F, message=F,warning=F,fig.align = "center",fig.width=10,fig.height=10}

drought_plot_data<- mydata %>% dplyr::select(ETa_ETp_0,drought_lag,year_fact) %>%
  gather(groups,measurements,ETa_ETp_0,drought_lag)

drought_plot<-ggplot(drought_plot_data,aes(y=measurements,x=as.factor(as.character(year_fact)),colour=groups)) +
  scale_x_discrete(breaks = c(1985:2022,1))+
  geom_boxplot(position=position_dodge(width=0.8))+
   geom_smooth(method = "gam", se=TRUE,aes(group=groups, colour=groups, fill=groups))+
  scale_colour_manual(name="", values=c("brown4","black"), labels=c("Drought (ETa/ETp) lag 4 years","Drought (ETa/ETp) no lag"))+
  scale_fill_manual(name="", values=c( "brown4","black"), labels=c("Drought (ETa/ETp) lag 4 years","Drought (ETa/ETp) no lag"))+
  
  labs(title="",y= "Drought indicator ETa/ETp", x = "")+
  theme_bw()+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
  theme(legend.position = c(0.15,0.1), axis.title.x=element_text(),axis.text.x=element_text(angle = 45, hjust = 1),legend.title=element_blank())
drought_plot
```



# Calculation of lag drought indicator ETa/ETp
```{r drought, echo = T, message=F,warning=F,fig.align = "center"}
# Bayesian multilevel binomial lag effect model using orthogonal polynomial

mydata <- mydata %>% 
  mutate(u0 = (ETa_ETp_0 + ETa_ETp_1 + ETa_ETp_2 + ETa_ETp_3 + ETa_ETp_4+ ETa_ETp_5), #mean
         u1 = ( 0*ETa_ETp_0 + 1*ETa_ETp_1 + 2*ETa_ETp_2 + 3*ETa_ETp_3 + 4*ETa_ETp_4 + 
                  5*ETa_ETp_5), #linear (to see if older years are more important, 
         # weighting of older years is higher in a linear way)
         u2 = ( 0*ETa_ETp_0 + 1*ETa_ETp_1 + 4*ETa_ETp_2 + 9*ETa_ETp_3 + 16*ETa_ETp_4+ 
                  25*ETa_ETp_5), #quadratic
         u3 = ( 0*ETa_ETp_0 + 1*ETa_ETp_1 + 8*ETa_ETp_2 + 27*ETa_ETp_3 + 64*ETa_ETp_4+ 
                  125*ETa_ETp_5)) #cubic

```
Note models are not fitted during knitting the pdf, due to time issues. The time for fitting one model can be up to 2 h, depending on the computer power. However the code is given for fitting the models, as # comments.

```{r drought2, echo = T, message=F,warning=F,fig.align = "center"}
# brms_poly_lags_drought <- brm( trees_died | trials(trees_died + trees_living)  ~
#                             
#                          u0 + u1 + u2 + u3 +
#                             
#                           (1|sites),
#                           data = mydata,
#                           family=zero_inflated_binomial(link = "logit"))

# saveRDS(brms_poly_lags_drought,
#         file = "models/brms_poly_lags_drought")
 #read it in again
brms_poly_lags_drought<-readRDS(file = "models/brms_poly_lags_drought_new") # 


# without u3
# brms_poly_lags_drought_1 <- brm( trees_died | trials(trees_died + trees_living)  ~
#                                  
#                                  u0 + u1 + u2 +
#                                  
#                                  (1|sites),
#                                data = mydata,
#                                family=zero_inflated_binomial(link = "logit"))
# 
# saveRDS(brms_poly_lags_drought_1,
#         file = "models/brms_poly_lags_drought_1")
#read it in again
brms_poly_lags_drought_1<-readRDS(file = "models/brms_poly_lags_drought_1_new") # 


# without u2
# brms_poly_lags_drought_2 <- brm( trees_died | trials(trees_died + trees_living)  ~
#                                    
#                                    u0 + u1 +
#                                  
#                                  (1|sites),
#                                  data = mydata,
#                                  family=zero_inflated_binomial(link = "logit"))
# 
# saveRDS(brms_poly_lags_drought_2,
#         file = "models/brms_poly_lags_drought_2")
#read it in again
brms_poly_lags_drought_2<-readRDS(file = "models/brms_poly_lags_drought_2_new") # 


# without u1
# brms_poly_lags_drought_3 <- brm( trees_died | trials(trees_died + trees_living)  ~
#                                    
#                                    u0 + 
#                                  
#                                  (1|sites),
#                                  data = mydata,
#                                  family=zero_inflated_binomial(link = "logit"))
# 
# saveRDS(brms_poly_lags_drought_3,
#         file = "models/brms_poly_lags_drought_3")
#read it in again
brms_poly_lags_drought_3<-readRDS(file = "models/brms_poly_lags_drought_3_new") # 


# Model comparison: Bayesian model evaluation using leave-one-out cross-validation (loo) 
# see Vehtari 2017.
 
# loo_compare(loo(brms_poly_lags_drought),loo(brms_poly_lags_drought_1),
# loo(brms_poly_lags_drought_2), loo(brms_poly_lags_drought_3))

# elpd_diff se_diff
# brms_poly_lags_drought_1   0.0       0.0
# brms_poly_lags_drought    -5.1       4.9
# brms_poly_lags_drought_3 -84.2      50.7
# brms_poly_lags_drought_2 -90.4      51.2

# brms_poly_lags_drought best model

# waic(brms_poly_lags_drought)
# # Estimate    SE
# # elpd_waic  -1383.7 126.6
# # p_waic       241.8  52.8
# # waic        2767.4 253.2


# waic(brms_poly_lags_drought_1)
# # Estimate    SE
# # elpd_waic  -1379.9 126.3
# # p_waic       235.7  50.4
# # waic        2759.8 252.7

# waic(brms_poly_lags_drought_2)
# # Estimate    SE
# # elpd_waic  -1466.4 132.9
# # p_waic       265.1  49.0
# # waic        2932.8 265.8


# waic(brms_poly_lags_drought_3)
# # Estimate    SE
# # elpd_waic  -1459.3 130.9
# # p_waic       255.4  46.8
# # waic        2918.6 261.8

# lagged effects plot of best poly lag model 
model <- brms_poly_lags_drought_1


summary(model)

# extract mean posterior distribution estimates of Bayesian model fit
coeff_poly_lag<-summary(model)$fixed
coeff_poly_lag <- coeff_poly_lag%>%
  dplyr::select(Estimate)

coeff_poly_lag<-as.data.frame(t(coeff_poly_lag))

### extraction of parameter estimates and their covariance matrix

# # extract mean posterior distribution estimates of Bayesian model fit
coeff_poly_lag<-summary(model)$fixed
coeff_poly_lag <- coeff_poly_lag%>%
  dplyr::select(Estimate)

(coeff_poly_lag<-as.data.frame(t(coeff_poly_lag)))

### extraction of parameter estimates and their covariance matrix

beta_1 <- coeff_poly_lag %>%
  dplyr::select(u0, u1, u2)

beta<-as.numeric(beta_1)

(c<-vcov(model, full = TRUE))#variance-coveariance matrix for a fitted model without intercept
#c_<-as.matrix(c[2:5, 2:5]) #matrix 4 u
 c_<-as.matrix(c[2:4, 2:4]) #matrix 3 u

c_

### generation of matrix for the computation of the effects at different lags

# # with 6 lags (lag 0- lag5) and 3 u (u0-u3):
(w<-rep(c(5,4,3,2,1,0),3)) #lag 0 - lag5  & u0, u1, u2, u3
(e<-c(rep(0,6),rep(1,6),rep(2,6))) #lag 0 - lag5  & u0, u1, u2, u3

(w_<-w^e)


 dim(w_)<-c(6,3) #lag 0 - lag5  & u0, u1, u2
w_
### computation of the lagged effects and of their covariance matrix
beta_<-w_%*%beta #  matrix multiplication
beta_ 
# or manual calculation without matrix multiplication
beta<-as.data.frame((beta_1))


lag_effects<-     as.data.frame(t(beta$u0+ (beta$u1*(0^1))+ (beta$u2*(0^2)) ))
lag_effects$lag1<-as.data.frame(t(beta$u0+ (beta$u1*(1^1))+ (beta$u2*(1^2)) ))
lag_effects$lag2<-as.data.frame(t(beta$u0+ (beta$u1*(2^1))+ (beta$u2*(2^2)) ))
lag_effects$lag3<-as.data.frame(t(beta$u0+ (beta$u1*(3^1))+ (beta$u2*(3^2)) ))
lag_effects$lag4<-as.data.frame(t(beta$u0+ (beta$u1*(4^1))+ (beta$u2*(4^2)) ))
lag_effects$lag5<-as.data.frame(t(beta$u0+ (beta$u1*(5^1))+ (beta$u2*(5^2)) ))

colnames(lag_effects)<-c("lag0","lag1","lag2","lag3","lag4","lag5")
lag_effects<-as.data.frame(t(lag_effects))
colnames(lag_effects)<-c("lag_effects")
lag_effects$lag<-rownames(lag_effects)

covb<-w_%*%c_%*%t(w_)
covb
### computation of the vector of standard errors of the lagged effects
se<-sqrt(diag(covb))

### Odds ratios associated with different lags, with 95%-confidence intervals
or<-exp(beta_)
ci95lo<-exp(beta_-1.96*se)
ci95up<-exp(beta_+1.96*se)
cbind(w_[,2],or,ci95lo,ci95up)

### If one wants to stay at the level of logits
ci95lo<-beta_-1.96*se
ci95up<-beta_+1.96*se
lag_effects_matrix<-cbind(w_[,2],beta_,ci95lo,ci95up)
lag_effects_matrix<-as.data.frame(lag_effects_matrix)
colnames(lag_effects_matrix)<-c("lag", "lag_effects", "ci95lo","ci95up" )
lag_effects_matrix$lag<-c("lag5","lag4","lag3","lag2","lag1","lag0")
lag_effects_matrix$lags<-c("lag 5 years","lag 4 years","lag 3 years","lag 2 years","lag 1 year","current year")

lag_effects_matrix <- lag_effects_matrix %>% mutate(lags = fct_reorder(lags,lag))# order




mydata <- mydata %>% 
  mutate(drought_lag_poly=(coeff_poly_lag$u0*u0)+(coeff_poly_lag$u1*u1)+(coeff_poly_lag$u2*u2)) 


hist(mydata$drought_lag_poly)

mydata$drought_lag_s<-scale(mydata$drought_lag_poly, center=TRUE, scale=TRUE)

```
\clearpage
## Lag effect plots

```{r lagplot, echo = T, message=F,warning=F,fig.align = "center",fig.width=5,fig.height=5}

lag_effects_matrix$signif<- c("no","no", "yes", "yes", "yes", "yes")
lag_effects_matrix<-lag_effects_matrix %>% filter(!lag=="lag5")

lag_effects_plot_drought<-ggplot(lag_effects_matrix,aes(x=lags,y=lag_effects)) + 
  geom_hline(yintercept=0, linetype="dotted", size=0.3) +
  geom_errorbar(aes(x=lags,y=lag_effects,ymin=ci95lo, ymax=ci95up), 
                width=.000001,size=0.5) +
  geom_point(aes(fill=signif),size=1.5, position = position_dodge(width = 0.8),
             pch=21,show.legend = F)+
  scale_fill_manual(values = c("white", "black"))+
  theme_bw()+ 
  ggtitle( TeX("Drought indicator ETa/ETp"))+ 
  coord_flip()+
  labs(x="", y="Estimates lag effects")+
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  theme(axis.ticks.x=element_line(size=0.1),axis.ticks.y=element_blank(),
        axis.line=element_line(),axis.line.x=element_line(size=0.1),
        axis.line.y=element_blank(),axis.title.x = element_text(size=4),
        axis.text.x  = element_text(size=4), 
        axis.text.y  = element_text(size=5),title = element_text(size=6))+
  theme(plot.title = element_text(hjust = 0.5))
lag_effects_plot_drought
```
\clearpage
# Correlation analysis
```{r corr_analysis, echo = T, message=F,warning=F,fig.align = "center"}

var<-mydata[,c( "drought_lag_s",
                "k_s","p_s","age_s","year_s","n_dep_s","bs_s")] #final model

sort(vif(var))
```
```{r corr_analysis2, echo = T, message=F,warning=F,fig.align = "center"}

source("Numerical_ecology_with_R_functions.R") #for cor.mtest function and ev plot
#rcorr Computes a matrix of Pearson's r rank correlation coefficients
res2 <- rcorr(as.matrix(var),type="pearson")
(p.mat <- cor.test.p(var))


```
Final variable selection for Bayesian zero-inflated binomial multilevel resgression: VIF: <1.8, r<=0.51 r< 0.65 in line with [@Brandl2020].



\clearpage
# Cross-correlation analysis crown damage vs. mortality
```{r CCV_analysis, echo = T, message=F,warning=F,fig.align = "center"}

# load crown defoliation data
crown_def<- read_delim("data/crown_def_data_spruce_2022.csv")
crown_def <- crown_def %>% mutate(year_fact=as.factor(as.character(year))) %>% 
  dplyr::select(-year)
mort_data <- mortality_rate %>% mutate(year_fact=as.factor(as.character(year_fact)))
mort_crow_data <- mort_data %>% left_join(crown_def,by="year_fact")
# passing to a long format
mort_crow_data<- mort_crow_data %>% mutate(date=as.POSIXct(year_fact, format = "%Y")) 
 
 long_tbl <- mort_crow_data%>%
  dplyr::select(date, 'Annual mean spruce crown damage > 25%'=mean_PROZ23, 
                'Annual mean spruce mortality'=mortality )%>%
  pivot_longer(cols = c('Annual mean spruce crown damage > 25%', 
                        'Annual mean spruce mortality'), names_to = "names", values_to = "value")
# the series with timetk package
time_series_plot<-long_tbl %>% 
  plot_time_series(.date_var = date,.value = value,.facet_vars = names, .title = "",
                   .x_lab = "", .y_lab = "Annual mean values (%)", .line_color = "black", 
                   .smooth_span=0.6, .smooth_color="grey")
time_series_plot

```
## CCF plot
```{r CCV_analysis2, echo = T, message=F,warning=F,fig.align = "center"}

# The cross correlation function is helpful for identifying lags of the x-variable 
# that might be useful predictors of yt


ccf(mort_crow_data$mean_PROZ23,mort_crow_data$mortality,
    main="",
    ylab = "CCF",
oma=c(2, 3, 5, 2))
mytitle = "Cross-correlation"
mysubtitle = "Annual mean spruce crown damage > 25% vs. Annual mean spruce mortality"
mtext(side=3, line=2, at=-12, adj=0, cex=1.5, mytitle)
mtext(side=3, line=1, at=-12, adj=0, cex=1, mysubtitle)
```

## Plot crown defoliation >25% vs. mortality
```{r CCV_analysis3, echo = T, message=F,warning=F,fig.align = "center"}


mort_crow_data<-mort_crow_data %>% 
    mutate(mean_PROZ23_lag_1 = dplyr::lag(mean_PROZ23,1),
           mean_PROZ23_lag_2 = dplyr::lag(mean_PROZ23,2),
           mean_PROZ23_lag_3 = dplyr::lag(mean_PROZ23,3),
           mean_PROZ23_lag_4 = dplyr::lag(mean_PROZ23,4))

# mean_PROZ23 lagged arithmetic mean
(mort_crow_data$mean_PROZ23_lag <- (mort_crow_data$mean_PROZ23 + mort_crow_data$mean_PROZ23_lag_1)/2)

#  spearman’s ranked correlation coefficient -> best correlation R=0.34,p=0.04
defoliation_plot<-ggplot(mort_crow_data, aes(y=mortality, x=mean_PROZ23_lag, colour=date)) +
  geom_point(size=2)+
     geom_smooth(method = "loess", span=0.9, colour="black",se = T, size=1)+
   scale_x_continuous(breaks = c(10,15,20,25))+
  scale_y_continuous(breaks = c(0,1,2,3,4,5))+
  geom_text(aes(label=year_fact),hjust=-0.3, vjust=-1, size=1.5)+
  annotate("text", x = 14, y = 5.2, size = 2.5,  colour="grey20",label = 
             expression(paste("Spearman's rank correlation coefficient\nR=0.36, p=0.032"))
           ,hjust = 1)+
  labs(title="",y= expression(paste("Annual mean spruce mortality rate (%)")), 
       x = "Lagged annual mean values of spruce with crown damages >25% (%)")+
  theme_bw()+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
  theme(legend.position = c(""),legend.title=element_blank())+
  coord_cartesian(ylim = c(0.1, 5.3)) # Zoom without cutting values
defoliation_plot
```

\clearpage



# Bayesian generalised multivariate multilevel binomial regression
The time for fitting one the Bayesian generalised multivariate multilevel binomial regression models can be several hours, depending on the computer power. However the code is given for fitting the models, as # comments.

```{r brms, echo = T, message=F,warning=F,fig.align = "center"}
# Note model selection is not shown here in order to save space (>200 lines of code 
# for comparing 25 differen models).
# brms_fit_final <- brm( trees_died | trials(trees_died + trees_living)  ~ # tree mortality
#                               drought_lag_s * n_dep_s + # drought indicator with 
# N deposition interaction
#                               bs_s +  # soil base saturation
#                               year_s + # missing time trend, not explained by other predictors
#                               age_s + # stand age
#                               p_s * n_dep_s + # foliar nutrients Phosphorus with 
# N deposition interaction 
#                               k_s  + # foliar nutrients Potassium
#                               (1|sites), # sites as random variable
#                             data = mydata,
#                             family=zero_inflated_binomial(link = "logit"), 
# chains = 4, iter=10000 )


#read it in again
brms_fit_final<-readRDS(file = "models/brms_fit_final_new_3") 
mod<-brms_fit_final
summary(mod)

```
## R-squared
R-squared for Bayesian regression models[@Gelman2019].
```{r brms2, echo = T, message=F,warning=F,fig.align = "center"}

# Bayesian version of R-squared for regression models, only fixed effects
bayes_R2(mod)  
```
\clearpage
## posterior predictive checks
Perform posterior predictive checks with the help of the bayesplot package [@Gabry2019].


```{r brms3, echo = T, message=F,warning=F,fig.align = "center"}

pp_check(mod, ndraws = 50)
```
## posterior distribution plot

```{r brms4, echo = T, message=F,warning=F,fig.align = "center"}

sum_table<-summary(mod)$fixed
sum_table<- sum_table %>% mutate(Bulk_ESS = as.character(round(Bulk_ESS,0)),
                                 Tail_ESS = as.character(round(Tail_ESS,0)))

colnames(sum_table) <- c ("Estimate" , "SD","Lower 95% CI" ,  "Upper 95% CI" , "Rhat" , 
                          "Bulk ESS" , "Tail ESS")
xtable(sum_table)

posterior_table_plot <- as.data.frame(fixef(mod)) %>%   droplevels() 
  
posterior_table_plot$fixed<-row.names(posterior_table_plot)

posterior_table_plot <-  posterior_table_plot %>% subset(!posterior_table_plot$fixed=="Intercept") %>%
  mutate(significant_negativ = ifelse(Estimate <0 & Q2.5 <0 & Q97.5 <0  ,"yes","no")) %>%
  mutate(significant_positiv = ifelse(Estimate>0 & Q2.5 >0 & Q97.5 >0  ,"yes","no")) %>%
  mutate(significant = ifelse(significant_negativ== "yes" | significant_positiv== "yes" ,"yes","no"))

# rename
 posterior_table_plot$fixed<-c("Drought lag","N deposition", "Base saturation","Year", 
                               "Stand age", "Phosphorus", "Potassium", "Drought lag:N deposition",
                               "Phosphorus:N deposition")

# reorder 
posterior_table_plot <- posterior_table_plot %>% mutate(fixed = fct_reorder(fixed, -Estimate))# order


coeff_plot<-ggplot(posterior_table_plot,aes(x=fixed,y=Estimate)) + 
  geom_hline(yintercept=0, linetype="dotted", size=0.2) +
  geom_errorbar(aes(x=fixed,y=Estimate,ymin=Q2.5 , ymax=Q97.5), width=.00001,size=0.5) +
  geom_point(size=2, position = position_dodge(width = 0.8), aes(shape=significant))+
  theme_bw()+ 
  scale_shape_manual(values=c(1,16))+
  scale_x_discrete(limits = rev(levels(posterior_table_plot$fixed)))+
  coord_flip()+
  labs(x="", y="Fixed effect estimates")+
  theme(panel.border = element_blank(), panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())+
  theme(axis.ticks.x=element_line(size=0.1),axis.ticks.y=element_blank(),
        axis.line=element_line(),axis.line.x=element_line(size=0.1),
        axis.line.y=element_blank(),axis.title.x = element_text(size=6),
        axis.text.x  = element_text(size=4), axis.text.y  = element_text(size=8),
        legend.position = "")
coeff_plot
```
\clearpage
## effect plots
### effect plot n_dep_s
```{r brms5, echo = T, message=F,warning=F,fig.align = "center"}

brms_pred<-conditional_effects(mod, "n_dep_s")$n_dep_s

brms_pred$effect1__<-(sd(mydata$n_dep)*brms_pred$effect1__)+mean(mydata$n_dep) 
# un-scaling z-transf

brms_effect_plot_n_dep<-ggplot(brms_pred,aes(x=effect1__,y=estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,show.legend = F,size=0.0001) +
     geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.48),
                                             labels = scales::percent_format(accuracy = 1))+
  labs( x = expression(paste("N deposition (kg N ", ha^-1, " ", a^-1, ")")),
        y = "Mortality Picea abies")+
  theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1), 
        axis.line=element_line(colour="black"),text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.text.x = element_text(size =12, colour = "black"),
        axis.line.x=element_line(size = 0.1),
        axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.1),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),
        legend.justification = c(1, 1), legend.position = c(1,1),
        legend.direction="horizontal",legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position = c(0.315,0.99),
        legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")
brms_effect_plot_n_dep
```

### effect plot drought_lag_s
```{r brms6, echo = T, message=F,warning=F,fig.align = "center"}
brms_pred<-conditional_effects(mod, "drought_lag_s")$drought_lag_s

brms_pred$effect1__<-(sd(mydata$drought_lag)*brms_pred$effect1__)+mean(mydata$drought_lag) 
# un-scaling z-transf

brms_effect_plot_drought_lag<-ggplot(brms_pred,aes(x=effect1__,y=estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,show.legend = F,size=0.0001) +
     geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.48),
                                             labels = scales::percent_format(accuracy = 1))+
  labs( x = "Drought (ETa/ETp)",y = "Mortality Picea abies")+
  theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1), 
        axis.line=element_line(colour="black"),text = element_text(size=14,colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.text.x = element_text(size =12, colour = "black"),
        axis.line.x=element_line(size = 0.1),axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.1),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),
        legend.background=element_blank(),
        legend.title=element_text(),
        legend.justification = c(1, 1), legend.position = c(1,1),
        legend.direction="horizontal",legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position = c(0.315,0.99),
        
        legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")+
  scale_x_reverse()
brms_effect_plot_drought_lag
```
### effect plot drought_lag_s:n_dep_s
```{r brms7, echo = T, message=F,warning=F,fig.align = "center"}

## unscaling N-deposition

nd <- -0.607 ## 20.00779
nd <- 0.477 ## 30.00437
nd <- 1.561## 40.00096

(sd(mydata$n_dep)*nd+mean(mydata$n_dep))

brms_pred<-conditional_effects(mod, "drought_lag_s:n_dep_s",
 int_conditions = list(n_dep_s=c( -0.607,0.477,1.561)))$drought_lag_s



brms_pred$effect1__<-(sd(mydata$drought_lag)*brms_pred$effect1__)+mean(mydata$drought_lag) 
# un-scaling z-transf

col.1<-"#d7191c" #red
col.2<-"#fdae61"
col.3<-"#abd9e9"
col.4<-"#2c7bb6" #blue

brms_effect_plot_drought_lag_s_n_dep_s<-ggplot(brms_pred,aes(x=effect1__,
                          y=estimate__,color=effect2__,fill=effect2__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,
              show.legend = F,size=0.0001) +
   geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.48),
                                           labels = scales::percent_format(accuracy = 1))+
  labs(y="Mortality Picea abies", x="Drought (ETa/ETp)",
       col=expression(atop("N deposition", paste("(kg N ",ha^-1, a^-1,")"))))+
  scale_fill_manual(values=c(col.1,col.2,col.4))+
  scale_colour_manual(values=c(col.1,col.2,col.4), label=c("40","30","20"))+
  theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1), 
        axis.line=element_line(colour="black"),text = element_text(size=14, 
        colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.text.x = element_text(size =12, colour = "black"),
        axis.line.x=element_line(size = 0.1),axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.1),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),legend.justification = c(1, 1), 
        legend.position = c(1,1),legend.direction="horizontal",
        legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position = c(0.315,0.99)
        ,legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")+
  scale_x_reverse( limits=c(1,0.7))                                             

brms_effect_plot_drought_lag_s_n_dep_s
```

### effect plot p_s:n_dep_s
```{r brms8, echo = T, message=F,warning=F,fig.align = "center"}
## unscaling N-deposition

nd <- -0.607 ## 20.00779
nd <- 0.477 ## 30.00437
nd <- 1.561## 40.00096

(sd(mydata$n_dep)*nd+mean(mydata$n_dep))

brms_pred<-conditional_effects(mod, "p_s:n_dep_s",
            int_conditions = list(n_dep_s=c( -0.607,0.477,1.561)))$p_s


brms_pred$effect1__<-(sd(mydata$p, na.rm=T)*brms_pred$effect1__)+
  mean(mydata$p, na.rm=T) # un-scaling z-transf

col.1<-"#d7191c" #red
col.2<-"#fdae61"
col.3<-"#abd9e9"
col.4<-"#2c7bb6" #blue

brms_effect_plot_p_s_n_dep_s<-ggplot(brms_pred,aes(x=effect1__,
                y=estimate__,color=effect2__,fill=effect2__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,
              show.legend = F,size=0.0001) +
   geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.48),
                labels = scales::percent_format(accuracy = 1))+
    scale_x_continuous( limits=c(0.59, 1.5))+
  geom_vline(aes(xintercept  =1.30), linetype="dashed", 
             size=0.5)+ #lower threshold for normal nutrition Gottlein2016
     labs(y="Mortality Picea abies", x=expression(paste(
       "Foliar phosphorus content (mg ",g^{-1}," dm)")),
       col=expression(atop("N deposition", paste("(kg N ",ha^-1, a^-1,")"))))+
  scale_fill_manual(values=c(col.1,col.2,col.4))+
  scale_colour_manual(values=c(col.1,col.2,col.4), 
                      label=c("40","30","20"))+
  theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1), 
  axis.line=element_line(colour="black"),text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
  axis.text.x = element_text(size =12, colour = "black"),
  axis.line.x=element_line(size = 0.1),
  axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", 
        fill=NA, size=0.1),panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),legend.justification = c(1, 1), 
        legend.position = c(1,1),legend.direction="horizontal",
        legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), 
        legend.position = c(0.315,0.99),
        legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")
brms_effect_plot_p_s_n_dep_s
```
### effect plot age_s
```{r brms9, echo = T, message=F,warning=F,fig.align = "center"}

brms_pred<-conditional_effects(mod, "age_s")$age_s

brms_pred$effect1__<-(sd(mydata$age)*brms_pred$effect1__)+mean(mydata$age) # un-scaling z-transf

brms_effect_plot_age<-ggplot(brms_pred,aes(x=effect1__,y=estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,show.legend = F,size=0.0001) +
    geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.48),
                                            labels = scales::percent_format(accuracy = 1))+
  labs( x="Stand age (a)",y = "Mortality Picea abies")+
  theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1), 
        axis.line=element_line(colour="black"),text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
  axis.text.x = element_text(size =12, colour = "black"),axis.line.x=element_line(size = 0.1),
  axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA,size=0.1),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),legend.justification = c(1, 1), 
        legend.position = c(1,1),legend.direction="horizontal",
        legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position = c(0.315,0.99),
        legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")
brms_effect_plot_age
```

### effect plot k_s
```{r brms10, echo = T, message=F,warning=F,fig.align = "center"}

brms_pred<-conditional_effects(mod, "k_s")$k_s

brms_pred$effect1__<-(sd(mydata$k,na.rm=T)*brms_pred$effect1__)+mean(mydata$k,na.rm=T) # un-scaling z-transf

brms_effect_plot_k_s<-ggplot(brms_pred,aes(x=effect1__,y=estimate__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,show.legend = F,size=0.0001) +
  geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.048),
            labels = scales::percent_format(accuracy = 1))+
  geom_vline(aes(xintercept  =4.5), linetype="dashed", size=0.5)+ #lower threshold for normal nutrition Gottlein2016
  labs(x=expression(paste("Foliar potassium concentration (mg ",g^{-1}," dm)")),
       y = "Mortality Picea abies")+
  theme(axis.title=element_text(size=12),
        axis.ticks = element_line(size=0.1), axis.line=element_line(colour="black"),text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
        axis.text.x = element_text(size =12, colour = "black"),axis.line.x=element_line(size = 0.1),
        axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.1),panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),legend.justification = c(1, 1), 
        legend.position = c(1,1),legend.direction="horizontal",
        legend.key.height=unit(1,"line"),plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(1, 1), legend.position = c(0.315,0.99),legend.direction="vertical")+
  theme(legend.title=element_text(size=10),legend.box.just = "left")
brms_effect_plot_k_s
```

# Bayesian generalised multivariate multilevel binomial regression altitude
The time for fitting one the Bayesian generalised multivariate multilevel binomial regression models can be several hours, depending on the computer power. However the code is given for fitting the models, as # comments.

```{r alti, echo = T, message=F,warning=F,fig.align = "center"}
# model with altitude
# # with zero_inflated_binomial family, only altitude ca. 2h
#  brms_fit_mod_alti <- brm( trees_died | trials(trees_died + trees_living)  ~
# 
#                           altitude_s +
# 
#                           (1|sites),
#                         data = mydata,
#                         family=zero_inflated_binomial(link = "logit"), chains = 4, iter=10000)
#  
#  saveRDS(brms_fit_mod_alti,
#          file = "models/brms_fit_mod_alti")



# read it in again
 brms_fit_mod_alti<-readRDS(file = "models/brms_fit_mod_alti") # 
 mod<-brms_fit_mod_alti
summary(mod)


```
## R-squared
R-squared for Bayesian regression models[@Gelman2019].
```{r alti2, echo = T, message=F,warning=F,fig.align = "center"}

# Bayesian version of R-squared for regression models, only fixed effects
bayes_R2(mod)  
```
\clearpage
## posterior predictive checks
Perform posterior predictive checks with the help of the bayesplot package [@Gabry2019].


```{r alti3, echo = T, message=F,warning=F,fig.align = "center"}

pp_check(mod, ndraws = 50)
```
## posterior distribution plot

```{r alti4, echo = T, message=F,warning=F,fig.align = "center"}

sum_table<-summary(mod)$fixed
sum_table<- sum_table %>% mutate(Bulk_ESS = as.character(round(Bulk_ESS,0)),
                                 Tail_ESS = as.character(round(Tail_ESS,0)))

colnames(sum_table) <- c ("Estimate" , "SD","Lower 95% CI" ,  "Upper 95% CI" , "Rhat" , 
                          "Bulk ESS" , "Tail ESS")
xtable(sum_table)

posterior_table_plot <- as.data.frame(fixef(mod)) %>%   droplevels() 
  
posterior_table_plot$fixed<-row.names(posterior_table_plot)

posterior_table_plot <-  posterior_table_plot %>% subset(!posterior_table_plot$fixed=="Intercept") %>%
  mutate(significant_negativ = ifelse(Estimate <0 & Q2.5 <0 & Q97.5 <0  ,"yes","no")) %>%
  mutate(significant_positiv = ifelse(Estimate>0 & Q2.5 >0 & Q97.5 >0  ,"yes","no")) %>%
  mutate(significant = ifelse(significant_negativ== "yes" | significant_positiv== "yes" ,"yes","no"))

# rename
 posterior_table_plot$fixed<-c("altitude")
 # reorder 
 posterior_table_plot <- posterior_table_plot %>% 
   mutate(fixed = fct_reorder(fixed, -Estimate))# order

 coeff_plot<-ggplot(posterior_table_plot,aes(x=fixed,y=Estimate)) + 
   geom_hline(yintercept=0, linetype="dotted", size=0.2) +
   geom_errorbar(aes(x=fixed,y=Estimate,ymin=Q2.5 , ymax=Q97.5), width=.00001,size=0.5) +
   geom_point(size=2, position = position_dodge(width = 0.8), 
              aes(shape=significant))+
   theme_bw()+ 
   scale_shape_manual(values=c(16,1))+
   scale_y_continuous(limits = c(-2,2))+
   scale_x_discrete(limits = rev(levels(posterior_table_plot$fixed)))+
   coord_flip()+
   labs(x="", y="Effects on the logit of mortality")+
   theme(panel.border = element_blank(), 
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank())+
   theme(axis.ticks.x=element_line(size=0.1),axis.ticks.y=element_blank(),
         axis.line=element_line(),axis.line.x=element_line(size=0.1),
         axis.line.y=element_blank(),axis.title.x = element_text(size=6),
         axis.text.x  = element_text(size=4), 
         axis.text.y  = element_text(size=8),
         legend.position = "")
 coeff_plot
 
```
```{r effectalti, echo = T, message=F,warning=F,fig.align = "center"}

 #effect plot altitude_s

 brms_pred<-conditional_effects(brms_fit_mod_alti, "altitude_s")$altitude_s
 
 brms_pred$effect1__<-(sd(mydata$altitude,na.rm=T)*brms_pred$effect1__)+
   mean(mydata$altitude,na.rm=T) # un-scaling z-transf
 
 brms_effect_plot_altitude<-ggplot(brms_pred,aes(x=effect1__,y=estimate__))+
   geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,
               show.legend = F,size=0.0001) +
   geom_line(size=1.5)+scale_y_continuous( limits=c(0,0.07),
            labels = scales::percent_format(accuracy = 1))+
   labs( x="Altitude (m.a.s.l.)",y = "Mortality Picea abies")+
   theme(axis.title=element_text(size=12),
         axis.ticks = element_line(size=0.1),
         axis.line=element_line(colour="black"),
         text = element_text(size=14, colour = "black"),
         axis.text.y = element_text(size = 12, colour = "black"),
         axis.text.x = element_text(size =12, colour = "black"),
         axis.line.x=element_line(size = 0.1),
         axis.line.y=element_line(size = 0.1)) +
   theme(panel.background=element_blank(),
         panel.border = element_rect(colour = "black", fill=NA, size=0.1),
         panel.grid.minor=element_blank(),
         panel.grid.major=element_blank(),
         legend.key=element_blank(),legend.background=element_blank(),
         legend.title=element_text(),legend.justification = c(1, 1), 
         legend.position = c(1,1),
         legend.direction="horizontal",legend.key.height=unit(1,"line"),
         plot.title = element_text(hjust = 0.5))+
   theme(legend.justification = c(1, 1), 
         legend.position = c(0.315,0.99),legend.direction="vertical")+
   theme(legend.title=element_text(size=10),legend.box.just = "left")+
   coord_cartesian(ylim = c(0,0.048)) # Zoom without cutting values
 brms_effect_plot_altitude
```
 
# Bayesian generalised multivariate multilevel binomial regression tree types
The time for fitting one the Bayesian generalised multivariate multilevel binomial regression models can be several hours, depending on the computer power. However the code is given for fitting the models, as # comments.

```{r tree, echo = T, message=F,warning=F,fig.align = "center"}
 # brms_fit_mod_tree_type <- brm( trees_died | trials(trees_died + trees_living)  ~
 #                                  tree_type  + # fixed effect tree type corrected for altitude
 #                                  (1|sites), # random effect sites
 #                                data = mydata,
 #                                family=zero_inflated_binomial(link = "logit"),
# chains = 4, iter=10000)
 # 
 # saveRDS(brms_fit_mod_tree_type,
 #         file = "models/brms_fit_mod_tree_type")



# read it in again
 brms_fit_mod_alti<-readRDS(file = "models/brms_fit_mod_tree_type") # 
 mod<-brms_fit_mod_alti
summary(mod)


```
## R-squared
R-squared for Bayesian regression models[@Gelman2019].
```{r tree2, echo = T, message=F,warning=F,fig.align = "center"}

# Bayesian version of R-squared for regression models, only fixed effects
bayes_R2(mod)  
```
\clearpage
## posterior predictive checks
Perform posterior predictive checks with the help of the bayesplot package [@Gabry2019].


```{r tree3, echo = T, message=F,warning=F,fig.align = "center"}

pp_check(mod, ndraws = 50)
```
## posterior distribution plot

```{r tree4, echo = T, message=F,warning=F,fig.align = "center"}

sum_table<-summary(mod)$fixed
sum_table<- sum_table %>% mutate(Bulk_ESS = as.character(round(Bulk_ESS,0)),
                                 Tail_ESS = as.character(round(Tail_ESS,0)))

colnames(sum_table) <- c ("Estimate" , "SD","Lower 95% CI" ,  "Upper 95% CI" , "Rhat" , 
                          "Bulk ESS" , "Tail ESS")
xtable(sum_table)

posterior_table_plot <- as.data.frame(fixef(mod)) %>%   droplevels() 
  
posterior_table_plot$fixed<-row.names(posterior_table_plot)

posterior_table_plot <-  posterior_table_plot %>% subset(!posterior_table_plot$fixed=="Intercept") %>%
  mutate(significant_negativ = ifelse(Estimate <0 & Q2.5 <0 & Q97.5 <0  ,"yes","no")) %>%
  mutate(significant_positiv = ifelse(Estimate>0 & Q2.5 >0 & Q97.5 >0  ,"yes","no")) %>%
  mutate(significant = ifelse(significant_negativ== "yes" | significant_positiv== "yes" ,"yes","no"))

 # rename
 posterior_table_plot$fixed<-c("tree type spruce")
 # reorder 
 posterior_table_plot <- posterior_table_plot %>% mutate(fixed = fct_reorder(fixed, -Estimate))# order
 
 coeff_plot<-ggplot(posterior_table_plot,aes(x=fixed,y=Estimate)) + 
   geom_hline(yintercept=0, linetype="dotted", size=0.2) +
   geom_errorbar(aes(x=fixed,y=Estimate,ymin=Q2.5 , ymax=Q97.5), width=.00001,size=0.5) +
   geom_point(size=2, position = position_dodge(width = 0.8), 
              aes(shape=significant))+
   theme_bw()+ 
   scale_shape_manual(values=c(16,1))+
   scale_y_continuous(limits = c(-2,2))+
   scale_x_discrete(limits = rev(levels(posterior_table_plot$fixed)))+
   coord_flip()+
   labs(x="", y="Effects on the logit of mortality")+
   theme(panel.border = element_blank(), panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
   theme(axis.ticks.x=element_line(size=0.1),
         axis.ticks.y=element_blank(),axis.line=element_line(),
         axis.line.x=element_line(size=0.1),axis.line.y=element_blank(),
         axis.title.x = element_text(size=6),
         axis.text.x  = element_text(size=4), 
         axis.text.y  = element_text(size=8),
         legend.position = "")
 coeff_plot

 
```
```{r effecttree, echo = T, message=F,warning=F,fig.align = "center"}

#effect plot 

 brms_pred<-conditional_effects(mod, "tree_type")$tree_type
 
 # est. error = sd
 #latex table output
 sum_table<-brms_pred
 sum_table<- sum_table %>% mutate(Estimate = as.character(round(estimate__*100,2)),
                                  SD = as.character(round(se__*100,2)),
                                  lower = as.character(round(lower__*100,2)),
                                  upper = as.character(round(upper__*100,2))) %>%
   
   dplyr::select(tree_type, Estimate,SD,lower,upper)
 
 colnames(sum_table) <- c ("Forest stands", "Estimate (%)" , "SD","Lower 95% CI" ,  "Upper 95% CI")
 xtable(sum_table)
 
 
 # brms_pred$effect1__<-(sd(mydata$n_dep)*brms_pred$effect1__)+mean(mydata$n_dep) # un-scaling z-transf
 col.1<-"#762a83" #purple
   col.2<-"#1b7837" #green
     
   brms_effect_tree_type <-ggplot(brms_pred,aes(x=effect1__,y=estimate__,colour=tree_type,
                                                shape=tree_type ))+
     geom_point(size=4)+
     geom_errorbar(width=0.05 ,aes(ymin =lower__,ymax =upper__ ))+
     scale_y_continuous( labels = scales::percent_format(accuracy = 1))+
     scale_x_discrete( label=c("mixed", "conifer"))+
     scale_colour_manual(values=c(col.1,col.2),label=c("mixed", "conifer"), name="Forest stands")+
     scale_shape_manual(values=c(16,15),label=c("mixed", "conifer"), name="Forest stands")+
     
     labs( x = expression(paste("Forest types")),y = "Mortality Picea abies")+
     theme(axis.title=element_text(size=12),axis.ticks = element_line(size=0.1),
           axis.line=element_line(colour="black"),
           text = element_text(size=14, colour = "black"),
        axis.text.y = element_text(size = 12, colour = "black"),
           axis.text.x = element_text(size =12, colour = "black"),
           axis.line.x=element_line(size = 0.1),
           axis.line.y=element_line(size = 0.1)) +
     theme(panel.background=element_blank(),
           panel.border = element_rect(colour = "black", fill=NA, size=0.1),
           panel.grid.minor=element_blank(),
           panel.grid.major=element_blank(),
           legend.key=element_blank(),legend.background=element_blank(),
           legend.title=element_text(),legend.justification = c(1, 1), 
           legend.position = c(1,1),
           legend.direction="horizontal",legend.key.height=unit(1,"line"),
           plot.title = element_text(hjust = 0.5))+
     theme(legend.justification = c(1, 1), legend.position = "c(0.315,0.99)",
           legend.direction="vertical")+
     theme(legend.title=element_text(size=10),legend.box.just = "left")+
     coord_cartesian(ylim = c(0,0.048)) # Zoom without cutting values
   brms_effect_tree_type
```
 

# Bayesian change-point regression model with brms
Concerning the applied method of hierarchical change-point regression models including random effects to estimate empirical critical loads for nitrogen using Bayesian Regression Models (brms) see [@Roth2022a].


The time for fitting one the Bayesian generalised multivariate multilevel binomial regression models can be several hours, depending on the computer power. However the code is given for fitting the models, as # comments.

```{r BCR, echo = T, message=F,warning=F,fig.align = "center"}

 ###6.1 CLempN estimation####

 # # binomial approch
 # # Full model approach with interaction without random intercepts
 # # CPR with interaction of drought
# mydata<- mydata%>% group_by(sites, year)%>%
#    mutate (n_trees= sum(trees_died,trees_living)) %>% ungroup()
#  mydata$Ndep <- mydata$n_dep #Time series data on total N deposition 
# were based on results of the EMEP MSC-W model (0.1° resolution) (EMEP, 2019). 
#  mydata$random<- as.integer(factor(mydata$sites)) #random effect sites
#  
#  mydata$fixeff1 <- mydata$p -0.94 # foliar Phosphorus concentrations
#  mydata$fixeff2 <- mydata$k -4.28 # foliar K concentrations
#  mydata$fixeff3 <-(sd(mydata$drought_lag_mean)*(mydata$drought_lag_poly*-1))+
# mean(mydata$drought_lag_mean) # un-scaling z-transf
#  mydata$fixeff3 <- mydata$fixeff3 -2.14
#  hist(mydata$fixeff3.1)
# hist(mydata$fixeff3)
#  
#  mydata$fixeff4 <- mydata$bs -56 # differences in soil base saturation
#  mydata$fixeff5 <- mydata$age -147  # stand age
#  mydata$fixeff6 <- mydata$year -2006  # unexplained year trend, not covered by other fixed effects
#  
#  
#  changepoint_model <- brmsformula(
#    trees_died | trials(n_trees) ~ beta0 + beta1 * fixeff1 + beta2 * fixeff2 + beta3 * fixeff3 + 
# beta4 * fixeff4 + beta5 * fixeff5 + beta6 * fixeff6 +
#      step(Ndep - CLempN) * betaN * (Ndep - CLempN)+ #steps of CLempN regression
#      betaNint * step(Ndep - CLempN) * (Ndep - CLempN)  * fixeff3, #interaction with drought
#    CLempN ~ 1+ (1|random), #group level effect 
#    beta0 ~ 1+ (1|random), #group level effect 
#    beta1 + betaN + betaNint + beta2 + beta3 + beta4+ beta5+ beta6 ~ 1,
#    nl = T,
#    family = binomial
#  )
#  
#   # Priors
#  bprior <- 
#    # prior(normal(0, 2), nlpar = "beta0") +
#    prior(normal(10, 5), nlpar = "beta0") +
#    prior(normal(10, 5), nlpar = "CLempN")+
#    prior(normal(0, 2),  nlpar = "betaN") + #vague priors
#    prior(normal(0, 2),  nlpar = "betaNint") + #vague priors
#    
#    prior(normal(0, 2),  nlpar = "beta1") + #vague priors
#    prior(normal(0, 2),  nlpar = "beta2") + #vague priors
#    prior(normal(0, 2),  nlpar = "beta3") + #vague priors
#    prior(normal(0, 2),  nlpar = "beta4") + #vague priors
#    prior(normal(0, 2),  nlpar = "beta5") + #vague priors
#    prior(normal(0, 2),  nlpar = "beta6")   #vague priors
#  
#  # Model fit with brms::brm using Stan
#  CPR_brms <- brms::brm(changepoint_model, 
#                        family = binomial,
#                        data = mydata, 
#                        prior = bprior, 
#                        chains = 2, 
#                        cores = 4, 
#                        warmup = 1000,
#                        thin = 2,
#                        iter=10000)
#  # # Save  object to file 
#  saveRDS(CPR_brms,
#          file = "models/CPR_brms_final_model_final")


CPR_brms<-readRDS(file = "models/CPR_brms_final_model_final")
 
summary(CPR_brms)
```
Rhat should be < 1.1 [@Koontz2021].

## R-squared
```{r BCR2, echo = T, message=F,warning=F,fig.align = "center"}


bayes_R2(CPR_brms) 
```
R-squared for Bayesian regression models according to [@Gelman2019]. 

## posterior predictive checks
Perform posterior predictive checks with the help of the bayesplot package [@Gabry2019].
```{r BCR3, echo = T, message=F,warning=F,fig.align = "center"}
pp_check(CPR_brms, ndraws = 50)

```
\clearpage

## effect plot interaction drought

```{r BCR5, echo = T, message=F,warning=F,fig.align = "center"}

brms_pred<- brms::conditional_effects(CPR_brms, "Ndep:fixeff3",
      int_conditions = list(fixeff3=c( 0.02,-0.03,-0.13)))$Ndep

col.1<-"#d7191c" #red
col.2<-"#fdae61"
col.3<-"#abd9e9"
col.4<-"#2c7bb6" #blue

brms_effect_plot_ClempN_drought_lag<-ggplot(brms_pred,aes(x=effect1__,y=estimate__,
                                                          color=effect2__,fill=effect2__))+
  geom_ribbon(aes(ymin = lower__, ymax = upper__), alpha = .1,
              show.legend = F,size=0.0001) +
  geom_line(size=1)+
  scale_y_continuous( labels = scales::percent_format(accuracy = 1))+
  scale_x_continuous( breaks=c(0,10,20,30,40,50,60,70,80))+
  
  geom_segment(aes(x = 10.86 , y = -1, xend = 10.86, yend = 0.6),linetype="solid",size=0.5, colour="grey20")+
  geom_segment(aes(x = 2.48, y = -1, xend = 2.48, yend = 0.6),linetype="dotted",size=0.5, colour="grey20")+
  geom_segment(aes(x = 19.05, y = -1, xend = 19.05, yend = .6),linetype="dotted",size=0.5, colour="grey20")+
  annotate("text", x = 19, y = 0.62, size = 2.8,  colour="grey20",
           label = expression(paste("Critical threshold N")),
           hjust = 1)+
  
  labs(y="Mortality Picea abies", x=expression(
    paste("N deposition (kg N ", ha^-1, " ", a^-1, ")")),
    col=expression("Drought (ETa/ETp)"))+
  scale_fill_manual(values=c(col.4,col.2,col.1))+
  scale_colour_manual(values=c(col.4,col.2,col.1), 
                      label=c("0.9","0.85","0.75"))+
  theme(axis.title=element_text(size=12),
        axis.ticks = element_line(size=0.1),
        axis.line=element_line(colour="black"),
        text = element_text(size=14, colour = "black"),
        
        axis.text.y = element_text(size = 12, 
                                   colour = "black"),
        axis.text.x = element_text(size =12, colour = "black"),axis.line.x=element_line(size = 0.1),
        axis.line.y=element_line(size = 0.1)) +
  theme(panel.background=element_blank(),
        panel.border = element_rect(colour = "black", fill=NA, size=0.1),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        legend.key=element_blank(),legend.background=element_blank(),
        legend.title=element_text(),
        legend.justification = c(1, 1), legend.position = c(1,1),legend.direction="horizontal",
        legend.key.height=unit(1,"line"),
        plot.title = element_text(hjust = 0.5))+
  theme(legend.justification = c(3.88, 1), legend.position = c(0.9998,0.99),legend.direction="vertical")+
  theme(legend.title=element_text(size=10),
        legend.box.just = "left")+

 coord_cartesian(ylim = c(-0.001, 0.8)) # Zoom without cutting values
brms_effect_plot_ClempN_drought_lag
```

\clearpage



# pinpoint method approach
The pinpoint method approach adapted according to [@Bobbink2010a].

```{r BCR6, echo = T, message=F,warning=F,fig.align = "center"}
 

mydata <- mydata %>%
  group_by(sites, year_fact) %>%
   mutate(mortality = 100*trees_died/(trees_died+trees_living)) %>%
  mutate(mortality = ifelse(is.na(mortality),0,mortality)) %>%
 ungroup()

mydata<-mydata %>%
  mutate(Ndep_class=case_when(n_dep < 15.01 ~ "<15",
                              n_dep >= 15 & n_dep < 20.01 ~ "15-20",
                              n_dep >= 20 & n_dep < 25.01 ~ "20-25",
                              n_dep >= 25 ~ ">25")) %>%
  mutate(Ndep_fact=as.factor(as.character(Ndep_class)))


# # reorder factor levels for plot
mydata<-mydata %>%
  mutate(Ndep_fact=factor(Ndep_fact, levels= c("<15","15-20","20-25",">25")))

pinpoint_plot<-ggplot(data=mydata %>% dplyr::filter(mortality>0))+
  geom_jitter(aes(x=Ndep_fact, y=mortality, color=year_fact),
  position = position_jitter(0.2),size=3) + 
  geom_boxplot(aes(x=Ndep_fact, y=mortality),alpha=0.1,outlier.shape = NA)+
  labs(y="Annual mean spruce mortality rate for sites with observed mortality (%)",
       x=expression(N~deposition~(kg~N~ha^-1~a^-1)))+
  ylim(0,10)+
  theme_bw()+
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())+
  theme(legend.position = c(0.09,0.88), axis.title.x=element_text(),
  axis.text.x=element_text(size=12),
  axis.ticks.x=element_blank(),legend.title=element_blank())

pinpoint_plot
```

\clearpage
## References
